% !TEX root = ../main.tex
\chapter*{Introduction}
\addcontentsline{toc}{chapter}{Introduction}
Many people were trying to discover what was in the beginning of the universe which we observe today.
Through the years, more or less successful theories were appearing and trying to describe its origin and behaviour.
Among them is one model, which provides a comprehensive explanation for a broad range of phenomena, including the cosmic microwave background, abundance of the light elements and Hubble's law.
This model is called The Big Bang theory and has born in 1927 on the basis of principles proposed by the Belgian priest and scientist Georges Lema{\^i}tre.
Using this model and known laws of physics one can calculate the characteristics of the universe in detail back in time to the extreme densities and temperatures.
However, at some point these calculations fail.
The extrapolation of the expansion of universe backwards in time using general relativity yields an infinite density and temperature at a finite time in the past.
This appearance of singularity is a signal of the breakdown of general relativity.
The range of this extrapolation towards singularity is debated - certainly we can go no closer than the end of \textit{Planck epoch} i.e. $10^{-43}$~s.
At this very first era the temperature of the universe was so high, that the four fundamental forces - electromagnetism, gravitation, weak nuclear interaction and strong nuclear interaction - were one fundamental force.
Between $10^{-43}$~s and $10^{-36}$~s of a lifetime of the universe, there is a \textit{grand unification epoch}, at which forces are starting to separate from each other.
The \textit{electroweak epoch} lasted from $10^{-36}$~s to $10^{-12}$~s, when the strong force separated from the electroweak force.
After the electroweak epoch, there was the \textit{quark epoch} in which the universe was a dense ``soup'' of quarks.
During this stage the fundamental forces of gravitation, electromagnetism, strong and weak interactions had taken their present forms. 
The temperature at this moment was still too high to allow quarks to bind together and form hadrons.
At the end of quark era, there was a big freeze-out - when the average energy of particle interactions had fallen below the binding energy of hadrons.
This era in which quarks became confined into hadrons is known as the hadron epoch.
At this moment the matter had started forming nuclei and atoms, which we observe today.

Here arises the question: how can we study the very beginning of the universe?
To do this, one should create in a laboratory a system with such a large density and high temperature to recreate those conditions.
Today, this is achievable through sophisticated machines, which are particle accelerators.
In the particle accelerators, like the Large Hadron Collider at CERN, Geneva or Relativistic Heavy Ion Collider at Brookhaven National Laboratory in Upton, New York, the heavy ions after being accelerated to near the speed of light are collided in order to generate extremely dense and hot phase of matter and recreate the quark-gluon plasma.
The plasma is believed to behave like an almost ideal fluid and to become a medium, that can be described by the laws of relativistic hydrodynamics.

This thesis is providing predictions for collective behaviour of the quark-gluon plasma coming from the hydrodynamic equations.
Experimental-like analysis was performed for the high energy Pb-Pb collisions generated with \verb|THERMINATOR| model.

The 1st chapter is an introduction to the theory of heavy ion collisions.
It contains the brief description of the Standard Model and Quantum Chromodynamics.
The quark-gluon plasma and its signatures are also characterized.

In the 2nd chapter there is a description of the relativistic hydrodynamic framework and the \verb|THERMINATOR| model used to perform the simulations of collisions.

The 3rd chapter covers the particle interferometry method used in this work.
The algorithm of building experimental correlation functions and effects coming from the hydrodynamics in the experimental results for particle interferometry are also presented.

In the 4th chapter there is a detailed analysis of the results for two-particle femtoscopy for different pairs of particles.
The quantitative analysis of calculated femtoscopic radii as well as the appearance of transverse mass scaling is discussed.